Distributional Trees for Circular Data

### 2019-02-01 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2423 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 734ae7b5450a9cbbef3a325e1bd648788ea7fdf5

% Brief abstract of the paper:
\abstract{Circular response variables occur in a variety of applications, e.g., when
geographical directions are considered or measurements are taken on a $24$-hour scale. 
A commonly chosen distribution for modeling circular data is the von Mises distribution.
As an alternative to estimating its distribution parameters by generalized additive models 
we propose the application of distributional trees which are capable of detecting also 
non-additive effects, select covariates and their possible interactions automatically and 
provide easily interpretable models allowing for a wide range of inference procedures. 
This approach is illustrated in an application to wind directions with a set of meteorological
covariates, all measured at the airport of Innsbruck, Austria.}

%which are capable of detecting also non-additive effects and provide easily interpretable 
%models allowing for a wide range of inference procedures.
%Moreover, covariates and their possible interactions are selected automatically
%such that no expert knowledge is required to specify the model in advance.

@@

%\subsection{Distributional Trees}
Distributional trees~(Schlosser et. al, 2018) embed distributional modeling into the framework 
of regression trees.
%Distributional trees~(Schlosser et. al, 2018) embed distributional modeling such as
%provided by generalized additive models for location, scale, and shape~(GAMLSS, 
%Rigby and Stasinopoulos, 2005) into the framework of regression trees~(Breiman et. al, 1984).
In particular one of the unbiased recursive partitioning algorithms MOB~(Zeileis et. al, 2008) 
and CTree~(Hothorn et. al, 2006) is applied and a distributional model is fitted in each node 
by estimating the parameters of a specified distribution family via maximum likelihood.
For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
tests evaluating associations between model scores and each possible split variable. After choosing 
the split variable showing the highest association to the goodness of fit of the model the corresponding split point is selected in a separate step by optimizing an objective function.

@@

In each node the distribution parameters $\mu$ and $\kappa$ are 
%vector $\theta = (\mu,\kappa)$ is 
estimated by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$ and model scores are calculated by 
$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \mu},
\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$. 
%$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
In that way a value indicating the goodness of fit is provided for each parameter and observation
such that changes can be detected not only in the location parameter $\mu$ but also in the 
dispersion parameter $\kappa$.

@@

%\begin{figure}
%\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
%\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
%\label{fig:tree}
%\end{figure}

@@

\section{Outlook}

This application shows that including covariates such as day time or earlier observations 
of wind directions as (unordered) factors provides valuable information. However, in order 
to properly account not only for the circularity of the response but also of regressors, 
additional adaptions of distributional trees are required. 
In particular, if the selected split variable is a circular variable this demands for 
finding a ``split interval'' instead of a split point.
%While the application of the von Mises distribution allows for modeling a circular response variable,
%additional adaptions have to be made when including circular regressor variables in a distributional tree
%model. In particular, if the selected split variable is a circular variable this demands for finding
%not only one split point but a ``split interval''. This consideration is of interest for many applications,
%for example when modeling wind directions it might be desirable to include day time or earlier 
%observations of wind directions as regressor variables.

Another possible extension of the here presented methodology is to build not just one tree 
model but an ensemble of trees to regularize and stabilize the model 
(see e.g. distributional forests, Schlosser et. al, 2018).

>>>>

% Brief abstract of the paper:
\abstract{
%Circular response variables occur in a variety of applications, e.g., when
%geographical directions are considered or measurements are taken on a $24$-hour scale. 
A commonly chosen distribution for modeling circular data is the von Mises distribution.
As an alternative to estimating its distribution parameters by generalized additive models 
we propose the application of distributional trees which are capable of detecting also 
non-additive effects, select covariates and their possible interactions automatically and 
provide easily interpretable models allowing for a wide range of inference procedures. 
This approach is illustrated in an application to wind directions with a set of meteorological
covariates, all measured at the airport of Innsbruck, Austria.}

@@

%\subsection{Distributional Trees}
Distributional trees~(Schlosser et. al, 2018) embed distributional modeling into the framework 
of regression trees.
%Distributional trees~(Schlosser et. al, 2018) embed distributional modeling such as
%provided by generalized additive models for location, scale, and shape~(GAMLSS, 
%Rigby and Stasinopoulos, 2005) into the framework of regression trees~(Breiman et. al, 1984).
The basic idea is to partition the covariate space recursively into subgroups 
such that a homogeneous distributional model can be fit to the response in each resulting subgroup.
In particular, one of the unbiased recursive partitioning algorithms MOB~(Zeileis et. al, 2008) 
and CTree~(Hothorn et. al, 2006) is applied and in each node the parameters of a specified 
distribution family are estimated via maximum likelihood.
For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
tests evaluating associations between model scores and each possible split variable. After choosing 
the split variable showing the highest association to the goodness of fit of the model the corresponding split point is selected in a separate step, either by optimizing the log-likelihood function (MOB)
or by evaluating a linear two-sample statistic (CTree) over all possible partitions.

@@

In each node of the tree the distribution parameters $\mu$ and $\kappa$ are 
%vector $\theta = (\mu,\kappa)$ is 
estimated by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$. 
For the resulting estimators $\hat{\mu}$ and $\hat{\kappa}$ model scores are calculated by 
$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \mu},
\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$
yielding a $n \times 2$-matrix for a data set of $n$ observations.
%$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
In that way a value indicating the goodness of fit is provided for each parameter and observation
such that changes can be detected not only in the location parameter $\mu$ but also in the 
dispersion parameter $\kappa$.

@@

\begin{figure}[!ht]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
\label{fig:tree}
\end{figure}

@@

\section{Discussion / Summary and Outlook}
Applying distributional trees on circular data provides easily interpretable models clearly
illustrating the impact of given covariates and their possible interactions on the distribution
of the dependent variable. Moreover, covariates are selected automatically such that no 
prespecification of the model is required which can be a challenging task, particularly for a 
meteorlogical quantity such as wind direction with high variation and many possible influencing 
factors.

This application shows that including covariates such as lagged observations of wind direction
as linear regressors provides valuable information. However, in order to properly account not 
only for the circularity of the response but also of regressors, additional adaptions of 
distributional trees are required. In particular, if the selected split variable is a circular 
variable this demands for finding a ``split interval'' instead of a split point.
This is also supported by the results illustrated in Figure~\ref{fig:tree} as mutliple splits in 
the linear regressor \textit{dd} (lagged wind direction) are required in order to represent its effect.

%While the application of the von Mises distribution allows for modeling a circular response variable,
%additional adaptions have to be made when including circular regressor variables in a distributional tree
%model. In particular, if the selected split variable is a circular variable this demands for finding
%not only one split point but a ``split interval''. This consideration is of interest for many applications,
%for example when modeling wind directions it might be desirable to include day time or earlier 
%observations of wind directions as regressor variables.

Another possible extension of the here presented methodology is to build not just one tree 
model but an ensemble of trees to regularize and stabilize the model 
(see e.g. distributional forests, Schlosser et. al, 2018). In that way, it should not have a 
significant impact on the results whether circular covariates are included as linear regressors
yielding a larger number of splits or as circular regressors with ``intervall splits''.
